---
title: 'Lab 2: Data wrangling'
output: html_document
---

In this first "real" lab we'll go over one of the most important skills in data science: Data Wrangling. It's the art of cleaning and manipulating our data to prepare it for our analyses and visualizations. A common estimate is that for most machine-learning projects, 80% of the time is spent on data wrangling, and the remaining 20% is spent on actually setting up and training the machine learning model.

Over the next two weeks, we'll do a lot of similar tasks, just in different ways. Today, we will focus on what is called Base R, which means using functions that are built into R, and we'll use a collection of functions in the package called `tidyverse`. The `tidyverse` contains tons of extremely useful functions that make data cleaning and manipulation a breeze. You might wonder then, why we aren't learning that right away. In R, you'll learn multiple ways to do things. Sometimes you'll read code that is written in a different way than you'd write it but it does the same thing, and it's good practice to get comfortable with the fact there are often multiple ways to do the same thing in R.

We'll start with basic data manipulation skills. It is useful to FIRST try to visualize in our head (or write down) what we want to do with the data, and THEN look for / think about the code that can do that for us. Once you learn some useful functions, you will be able to combine them to manipulate the data to your will like a wizard.


## Part 1: Base R

### Getting an overview of the data

First we'll explore different ways to get an overview of the dataset.
Let's use the `gapminder` dataset, available in the `dslabs` package. Because the data is available in the package, we can load it by simply using `data(gapminder)`. First we'll look at the dataset to see what it contains. We will use `summary()`, `head()` and `str()`.

```{r, message=FALSE}
# install.packages('dslabs') # Only run if you haven't installed dslabs
library(dslabs)
data(gapminder)
df <- gapminder
```

```{r}
summary(df)

head(df)

str(df)
```

1. What type of information do each one of these functions provide about our dataset?

* Your answer here (no code needed)

2. How is the dataset ordered?

* Your answer here (no code needed)

3. What range of years does our dataset contain?

* Your answer here (no code needed)


### Ordering and subsetting our dataset

We can subset our dataset in different ways. For example, the brute-force approach: we can select the cell in row 1, column 3 using `gapminder[1,3]`. We can also select a range of rows using `:`; i.e., we can select rows 2 to 5, column 1 using `gapminder[2:5,1]`.

```{r}
# Subsetting via numeric indexes

gapminder[1,3] # Select row 1, column 3

gapminder[2:5,1] # Select rows 2-5, column 3

gapminder[c(1,3,5), c(1,3,5)] # Select rows 1,3,5 columns 1,3,5

gapminder[1,] # entire first row, all columns (note we still need comma)

head(gapminder[,1]) # entire first column, all rows (note we still need comma)
```


4. In general, selecting rows and columns via numeric indices may not always be the best idea. What are some disadvantages you can think of?

* Your answer here (no code needed)

Another way to select columns is with a vector of column names:

```{r}
# Select a subset of columns with a character vector
head(gapminder[, c('country', 'year', 'infant_mortality')])
```


### Filtering rows with a logical

A logical is just a vector of TRUE and FALSE values. Let's look first at a simple example of a vector, before moving to our full dataset.

```{r}
# Save the numbers 1-10
a <- 1:10

# Logical comparisons
a < 3 # Values < 3 will be TRUE, otherwise FALSE

a <= 2 | a > 8 # Values <=2 OR > 8 will be TRUE, otherwise FALSE

a >= 3 & a <= 7 # Values >= 3 AND <= 7 will be TRUE, otherwise FALSE

# Subsetting a via logical comparisons
# Just returns the elements that are true
a[a < 3]
a[a <= 2 | a > 8]
a[a >= 3 & a <= 7]
```


Now let's go back to the `gapminder` dataset. Filter the dataset to only observations (rows) for the United States, and only the `country`, `year`, and `infant_mortality` columns.

```{r}
# Filter to just the United States
gapminder[gapminder$country == 'United States',]

# Combine Filter + Subset the columns
gapminder[gapminder$country == 'United States', c('country', 'year', 'infant_mortality')]
```

What happens with NAs? We see that `infant_mortality` has `NA` values:

```{r}
head(gapminder[gapminder$infant_mortality > 10,])
```

When we do this, we see that values where infant mortality are `NA` are kept. We can get around this by using the function `which()`. The `which()` function returns the index of rows in a data frame that meet a certain criteria. For example, if we want to know the index of the rows for Algeria, we can type the following:

```{r}
which(gapminder$country == "Algeria")
```

We can use this to select rows within our data frame by using square brackets `[ ]` and placing our `which()` function where we would place the row number.

```{r}
head(gapminder[which(gapminder$country == "Algeria"), ])

head(gapminder[which(gapminder$country == "Algeria"), "infant_mortality"])
```

Now, let's use this to filter rows without NA values.

```{r}
ind <- which(gapminder$infant_mortality > 10)
head(gapminder[ind,])
```


5. Create a data frame called `algeriagdp` that includes only year and GDP data for Algeria.

```{r}
# Your code here
```


### Ordering Data

The function order returns the numeric indices needed to sort the data.

```{r}
a <- c(1,3,5,2,4,8)
b <- c('apple', 'orange', 'banana', 'pear', 'grape', 'lemon')

# Sort a
sort(a)

order(a) # order in which to get the elements of a so that it is sorted
a[order(a)] # equivalent to sort
b[order(a)] # sort b by a
```

Let's try it on the `gapminder` dataset.

```{r}
# Application to the data frame
head(gapminder[order(gapminder$infant_mortality),])
head(gapminder[order(gapminder$life_expectancy, decreasing = T),])
```

6. Create a data frame called `df2` that is ordered so that it shows the data for each country (all years), then the next country (all years), etc.

```{r}
# Your code here
```


### Creating new data

Lastly, we will look at creating new data or editing data.

7. Create a new column of GDP per capita.

```{r}
# Your code here
```

8. Transform `country` from a factor variable to a character variable. Then rename `United States` to `USA`.

```{r}
# Your code here
```


---

## Part 2: Tidyverse

Now, we will do the same types of operations but using the tidyverse functions.

We can use the `filter()` function to select rows according to values of a column, and we can use `select()` to select only specific columns according to their name. If we want a specific set of rows, we can use `slice()`.

```{r, warning = F, message = F}
library(dplyr)
```

First, let's use `filter()`. Notice that we don't need to use `df$` in front of the names, since we supply `df` as the first argument to the function.

```{r}
# Filter
head(filter(df, country == 'United States'))
head(df[df$country == 'United States',])

# Notice that filter automatically throws out NA -- no need for which()!
head(filter(df, infant_mortality < 10))
head(df[df$infant_mortality < 10,])

# Multiple combinations
head(filter(df, infant_mortality < 10 | life_expectancy > 80))
```

1. How many rows have `infant_mortality > 20` and `life_expectancy > 70`?

```{r}
# Your code here
```


Next we can use `select()` to get column names. `select()` doesn't require that we use quotes around names.

```{r}
# Select 3 columns (quotes work fine but not required)
head(select(df, country, year, infant_mortality))
head(select(df, 'country', 'year', 'infant_mortality'))

# Select everything except the region column
head(select(df, -region))
```

```{r}
### 1st row of df
slice(df, 1)

### rows 20-50 
slice(df, 20:50)
```


We can also order a data frame using the `arrange()` function.

```{r}
# Default is increasing
head(arrange(df, infant_mortality))
head(df[order(df$infant_mortality),]) # this is a little ugly

# Use desc() for decreasing
head(arrange(df, desc(infant_mortality)))
head(df[order(df$infant_mortality, decreasing = T),]) # even more ugly

# Sort by continent, then region
head(arrange(df, continent, region))
head(df[order(df$continent, df$region),]) # even more ugly


# First sort by continent [A-Z] then in descending order of infant mortality rate
df4 <- arrange(df, continent, desc(infant_mortality)) # much prettier/simpler!
```

2. Create a dataset that includes country, year, gdp and population for countries in South-Eastern Asia between 1965 and 2010. Order the data frame by country, then year. Confirm that your code is doing what it's supposed to.

```{r}
# Your code here
```


### The pipe operator `%>%`

To generate the result above, you may have needed multiple steps or lines of code. We can use the pipe operator `%>%` to avoid using some of the intermediate steps and to cut down on the amount of code we need to type. This makes for much more intuitive and fun data wrangling.

For example, we can create the data frame mentioned above in an elegant, intuitive way:

```{r}
algeriagdp <- gapminder %>% 
  filter(country == 'Algeria') %>% 
  select(gdp, year)

# One way
x <- filter(gapminder, country == 'Algeria')
head(select(x, gdp, year))

# Another way (hard to read)
head(select(filter(gapminder, country == 'Algeria'), gdp, year))

```

3. Now, using the pipe operator, create a dataset that includes `country`, `year`, `gdp` and `population` for countries in South-Eastern Asia between 1965 and 2010. Order the data frame by `country`, then `year`. Try to do it all in one command, starting from the `gapminder` dataset and using the pipe operator. Confirm that your code is doing what it's supposed to by printing out the result.

```{r}
# Your code here
```

### Creating variables

We can create variables using the function `mutate()`.

```{r}
gapminder <- mutate(gapminder, 'gdp_per_capita' = gdp/population)

# Same thing but using the pipe
gapminder <- gapminder %>% 
  mutate('gdp_per_capita' = gdp/population)
```

4. Create a variable with the approximate number of women in the population. Assume that 52% of the population are women.

```{r}
# Your code here
```

5. Calculate an approximation of the number of children born each year based on the `fertility` column, which displays the expected number of children per woman over their reproductive life. For this, assume an average reproductive life of 25 years.

```{r}
# Your code here
```

--- 

You can find a list of our spells... ehhhm, functions [here](https://rstudio.github.io/cheatsheets/data-transformation.pdf).


And a list of other useful spellbooks (cheat sheets) [here](https://posit.co/resources/cheatsheets/). 

