## Reshaping data

As we have seen, having data in `tidy` format is what makes the `tidyverse` flow. After the first step in the data analysis process, importing data, a common next step is to reshape the data into a form that facilitates the rest of the analysis. The `tidyr` package includes several functions that are useful for tidying data. 

We will use the fertility wide format dataset described earlier as an example in this section.

```{r, message=FALSE, warning=FALSE}
library(tidyverse) 
library(dslabs)
filename <- "https://raw.githubusercontent.com/rafalab/dslabs/master/inst/extdata/fertility-two-countries-example.csv"
wide_data <- read_csv(filename)
head(wide_data)
```

### `pivot_longer`

One of the most used functions in the tidyr package is `pivot_longer`, which is useful for converting wide data into tidy data.

As with most tidyverse functions, the `pivot_longer` function's first argument is the data frame that will be converted. Here we want to reshape the `wide_data` dataset so that each row represents a fertility observation, which implies we need three columns to store the year, country, and the observed value. In its current form, data from different years are in different columns with the year values stored in the column names. Through the `names_to` and `values_to` argument we will tell `pivot_longer` the column names we want to assign to the columns containing the current column names and observations, respectively. The default names are `name` and `value`, which are often usable choices. In this case a better choice for these two arguments would be `year` and `fertility`. Note that nowhere in the data file does it tell us this is fertility data. Instead, we deciphered this from the file name. Through `cols`,the second argument we specify the columns containing observed values; these are the columns that will be pivoted. The default is to pivot all columns so, in most cases, we have to specify the columns. In our example we want columns `1960`, `1961` up to `2015`.

The code to pivot the fertility data therefore looks like this:

```{r}
new_tidy_data <- pivot_longer(wide_data, `1960`:`2015`, names_to = "year", values_to = "fertility")
```

We can also use the pipe like this:

```{r}
new_tidy_data <- wide_data %>% 
  pivot_longer(`1960`:`2015`, names_to = "year", values_to = "fertility")
```

We can see that the data have been converted to tidy format with columns `year` and `fertility`:

```{r}
head(new_tidy_data)
```

and that each year resulted in two rows since we have two countries and this column was not pivoted. A somewhat quicker way to write this code is to specify which column will not include in the pivot, rather than all the columns that will be pivoted:

```{r}
new_tidy_data <- wide_data %>%
  pivot_longer(-country, names_to = "year", values_to = "fertility")
```

The `new_tidy_data` object looks like the original `tidy_data` we defined this way

```{r}
data("gapminder")
tidy_data <- gapminder %>% 
  filter(country %in% c("South Korea", "Germany") & !is.na(fertility)) %>%
  select(country, year, fertility)
```

with just one minor difference. Can you spot it? Look at the data type of the year column:

```{r}
class(tidy_data$year)

class(new_tidy_data$year)
```

The `pivot_longer` function assumes that column names are characters. So we need a bit more wrangling before we are ready to make a plot. We need to convert the year column to be numbers:

```{r}
new_tidy_data <- wide_data %>%
  pivot_longer(-country, names_to = "year", values_to = "fertility") %>%
  mutate(year = as.integer(year))
```

Note that we could have also used the `mutate` and `as.numeric`.

Now that the data is tidy, we can use this relatively simple ggplot code:

```{r}
new_tidy_data %>% ggplot(aes(year, fertility, color = country)) + 
  geom_point()
```


### `pivot_wider`

As we will see in later examples, it is sometimes useful for data wrangling purposes to convert tidy data into wide data. We often use this as an intermediate step in tidying up data. The `pivot_wider` function is basically the inverse of `pivot_longer`. The first argument is for the data, but since we are using the pipe, we don’t show it. The `names_from` argument tells `pivot_wider` which variable will be used as the column names. The `values_from` argument specifies which variable to use to fill out the cells.

```{r}
new_wide_data <- new_tidy_data %>% 
  pivot_wider(names_from = year, values_from = fertility)
select(new_wide_data, country, `1960`:`1967`)
```

Similar to `pivot_longer`, `names_from` and `values_from` default to name and value.


### `separate`

The data wrangling shown above was simple compared to what is usually required. In our example spreadsheet files we include an example that is slightly more complicated. It includes two variables: life expectancy as well as fertility. However, the way it is stored is not tidy and, as we will explain, not optimal.

```{r, message=FALSE}
path     <- system.file("extdata", package = "dslabs")
filename <- file.path(path, "life-expectancy-and-fertility-two-countries-example.csv")

raw_dat <- read_csv(filename)
select(raw_dat, 1:5)
```

First, note that the data is in wide format. Second, notice that this table includes values for two variables, fertility and life expectancy, with the column name encoding which column represents which variable. Encoding information in the column names is not recommended but, unfortunately, it is quite common. We will put our wrangling skills to work to extract this information and store it in a tidy fashion.

We can start the data wrangling with the `pivot_longer function`, but we should no longer use the column name year for the new column since it also contains the variable type. We will call it `name`, the default, for now:

```{r}
dat <- raw_dat %>% pivot_longer(-country)
head(dat)
```

The result is not exactly what we refer to as tidy since each observation is associated with two, not one, rows. We want to have the values from the two variables, fertility and life expectancy, in two separate columns. The first challenge to achieve this is to separate the `name` column into the year and the variable type. Notice that the entries in this column separate the year from the variable name with an underscore: 

```{r}
dat$name[1:5]
```

Encoding multiple variables in a column name is such a common problem that the `readr` package includes a function to separate these columns into two or more. Apart from the data, the `separate` function takes three arguments: the name of the column to be separated, the names to be used for the new columns, and the character that separates the variables. So, a first attempt at this is:

```{r, eval=FALSE}
dat %>% separate(name, c("year", "variable_name"), "_")
```

Because `_` is the default separator assumed by `separate`, we do not have to include it in the code:

```{r}
dat %>% separate(name, c("year", "variable_name"))
```

The function does separate the values, but we run into a new problem. We receive the warning `Too many values at 112 locations:` and that the `life_expectancy` variable is truncated to life. This is because the `_` is used to separate `life` and `expectancy`, not just year and variable name! We could add a third column to catch this and let the `separate` function know which column to fill in with missing values, `NA`, when there is no third value. Here we tell it to fill the column on the right:

```{r}
var_names <- c("year", "first_variable_name", "second_variable_name")
dat %>% separate(name, var_names, fill = "right")
```

However, if we read the `separate` help file, we find that a better approach is to merge the last two variables when there is an extra separation:

```{r}
dat %>% separate(name, c("year", "name"), extra = "merge")
```

This achieves the separation we wanted. However, we are not done yet. We need to create a column for each variable. As we learned, the `pivot_wider` function can do this:

```{r}
dat %>% 
	separate(name, c("year", "name"), extra = "merge") %>%
  			pivot_wider() 
```

The data is now in tidy format with one row for each observation with three variables: `year`, `fertility` and `life expectancy`.

### `unite`

It is sometimes useful to do the inverse of `separate`, i.e. unite two columns into one. To demonstrate how to use unite, we show code that, although not the optimal approach, serves as an illustration. Suppose that we did not know about extra and used this command to separate: 

```{r}
var_names <- c("year", "first_variable_name", "second_variable_name")
dat %>% 
  separate(name, var_names, fill = "right")
```

we can achieve the same final result by uniting the second and third columns, then pivoting the columns and renaming `fertility_NA` to `fertility`:

```{r}
dat %>% 
  separate(name, var_names, fill = "right") %>%
  unite(name, first_variable_name, second_variable_name) %>%
  pivot_wider() %>%
  rename(fertility = fertility_NA)
```



